The algebraic hierarchy in type theory using type classes

### 2010-06-22 jar.tex
Improve presentation of Coq definitions of numerical interfaces.

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Class Naturals A {e plus mult zero one} `{!NaturalsToSemiRing A} :=
    { naturals_ring:> @SemiRing A e plus mult zero one
    ; naturals_to_semiring_mor:> Π `{SemiRing B}, SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (unary, binary, factor multisets, arrays of native machine words, etc.).

@@

The abstract interface for integers is completely analogous to the one for natural numbers:
\begin{lstlisting}
  Class Integers A {e plus mult zero one opp} `{IntegersToRing A} :=
    { integers_ring:> @Ring A e plus mult zero one opp
    ; integers_to_ring_mor:> Π `{Ring B}, Ring_Morphism (integers_to_ring A B)
    ; integers_initial:> Initial (ring.object A) }.
\end{lstlisting}

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and for which division of such integers is surjective:
\begin{lstlisting}
  Class Rationals A {e plus mult opp zero one mult_inv } `{Π x y: A, Decision (x = y)} {inj_inv}: Prop :=
    { rationals_field:> @Field A e plus mult opp zero one mult_inv
    ; rationals_frac: Surjective (λ p => integers_to_ring (Z nat) A (fst p) * / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinling|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinilne|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

>>>>

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (unary, binary, factor multisets, arrays of native machine words, etc.).

@@

The abstract interface for integers is completely analogous to the one for natural numbers:
\begin{lstlisting}
  Context `{Ring A} `{IntegersToRing A}.
  Class Integers: Prop :=
    { integers_ring:> Ring A
    ; integers_to_ring_mor:> Π `{Ring B},
        Ring_Morphism (integers_to_ring A B)
    ; integers_initial:> Initial (ring.object A) }.
\end{lstlisting}

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and for which division of such integers is surjective:
\begin{lstlisting}
  Context `{Field A} `{Π x y: A, Decision (x = y)} {inj_inv}.
  Class Rationals: Prop :=
    { rationals_field:> Field A
    ; rationals_frac: Surjective
        (λ p => integers_to_ring (Z nat) A (fst p) *
          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked any other (or even an explicitly arbitrary) implementation of \lstinline|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

% todo: say a word about section variable/context discharging, because we use it everywhere.
